---
title: "Exploration of Analyses for Poland COVID"
author: "Caulu Lin"
date: "2024-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r packages}
pacman::p_load(dplyr, ggplot2, ggstatsplot)
```


```{r data and test for salt }
## How much salt do you use in your meals? None	Very little	Little	Moderately	  Much	Very much
 
df<-data.frame(row.names =c("None", "VeryLittle", "Litlle", "Moderately", "Much", "VeryMuch"),
              CoV_smell_taste_disturbances=c(0,3,4,12,6,0),
              CoV_no_smell_taste_disturbances=c(0,2,1,3,1,0),
               healthCtr=c(0,8,7,18,6,0),
               pertCoV_smell_taste_disturbances=c(0,12,16,48,24,0),
               pertCoV_no_smell_taste_disturbances=c(0,28.6,14.3,42.9,14.3,0),
               perthealthCtr=c(0,20.5,17.9,46.2,15.4,0)) 


# CoV_smell_taste_disturbances vs healthCtr # Count
chisq.test(df[ c(1,3)])
chisq.test(df[ c(4,6)])

#
fisher.test(df[c(1,3)])
fisher.test(df[c(1,3)])
#
ks.test(df$CoV_smell_taste_disturbances, df$healthCtr)
ks.test(df$pertCoV_smell_taste_disturbances, df$perthealthCtr)

#

```



```{r data and test for sugar}
df<-data.frame(row.names =c("None", "VeryLittle", "Litlle", "Moderately", "Much", "VeryMuch"),
              CoV_smell_taste_disturbances=c(2,5,7,8,2,1),
              CoV_no_smell_taste_disturbances=c(0,3,1,2,1,0),
               healthCtr=c(2,6,10,12,9,0),
               pertCoV_smell_taste_disturbances=c(8,20,28,32,8,4),
               pertCoV_no_smell_taste_disturbances=c(0,42.9,14.3,28.6,14.3,0),
               perthealthCtr=c(5.1,15.4,25.6,30.8,23.1,0)) 


# CoV_smell_taste_disturbances vs healthCtr # Count
chisq.test(df[ c(1,3)])
chisq.test(df[ c(4,6)])

#
fisher.test(df[c(1,3)])
fisher.test(df[c(1,3)])
#
ks.test(df$CoV_smell_taste_disturbances, df$healthCtr)
ks.test(df$pertCoV_smell_taste_disturbances, df$perthealthCtr)

```
## we code data as numeric data as
## None=0, VeryLittle=1, Little=2, moderately=3, much=4, VeryMuch=5
```{r data coding and test}
df<-data.frame(group=c(rep("CoV_smell_taste_disturbances",25), rep("CoV_no_smell_taste_disturbances", 7), rep("healthCtr", 39)), salt= c(1,	1,	1,	2,	2,	2,	2,	3,	3,	3,	3,	3,	3,	3,	3,	3,	3,	3,	3,	4,	4,	4,	4,	4,	4,1,	1,	2,	3,	3,	3,	4,1,	1,	1,	1,	1,	1,	1,	1,	2,	2,	2,	2,	2,	2,	2,	3,	3,	3,	3,	3,	3,	3,	3,	3,	3,	3,	3,	3,3,	3,	3,	3,	3,	4,	4,	4,	4,	4,	4), sugar_sweeteners=c(0,	0,	1,	1,	1,	1,	1,	2,	2,	2,	2,	2,	2,	2,	3,	3,	3,	3,	3,	3,	3,	3,	4,	4,	5,	1,	1,	1,	2,	3,	3,	4,	0,	0,	1,	1,	1,	1,	1,	1,	2,	2,	2,	2,	2,	2,	2,	2,	2,	2,	3,	3,	3,	3,	3,	3,	3,	3,	3,	3,	3,	3,	4,	4,	4,	4,	4,	4,	4,	4,	4))


wilcox.test(salt ~ group,subset(df, group !="CoV_no_smell_taste_disturbances"))


ggbetweenstats(
  data = df,
  x = group,
  y = salt
)

wilcox.test(sugar_sweeteners ~ group,subset(df, group !="CoV_no_smell_taste_disturbances"))

ggbetweenstats(
  data = df,
  x = group,
  y = sugar_sweeteners
)
```

#COVID patients who experience a loss of taste or smell tend to increase their salt intake and reduce their sugar/sweetener intake, as expected. To determine the true difference between the two groups, it is important to adjust for sex and age as covariates. Including these covariates can strengthen the results by accounting for their potential influence on taste and smell, enabling the model to more accurately isolate the effects of the primary variables.

##Given the small sample size, advanced statistical methods and resampling techniques can provide more robust insights. 

# Bootstrap
```{r Bootstrap}
library(boot)

##
statistic_function <- function(data, indices) {
  df <- data[indices, ]
  return(mean(df$salt[df$group == "CoV_smell_taste_disturbances"]) - mean(df$salt[df$group == "healthCtr"]))
}

salt_bootstrap_results <- boot(data = df, statistic = statistic_function, R = 1000)
salt_bootstrap_results

##sugar_sweeteners
statistic_function <- function(data, indices) {
  df <- data[indices, ]
  return(mean(df$sugar_sweeteners[df$group == "CoV_smell_taste_disturbances"]) - mean(df$sugar_sweeteners[df$group == "healthCtr"]))
}

sugar_bootstrap_results <- boot(data = df, statistic = statistic_function, R = 1000)
sugar_bootstrap_results
```

# MCMC
```{r Markov Chain Monte Carlo (MCMC) used in Bayesian model}
library(rjags)
# Define a Bayesian model in JAGS syntax
model_string <- "model {
 # Likelihood
  for (i in 1:N) {
    salt[i] ~ dnorm(mu, tau)
  }

  # Priors
  mu ~ dnorm(0, 0.001)
  tau ~ dgamma(0.001, 0.001)
}
"
# Prepare data for JAGS
df1<-subset(df, group !="CoV_no_smell_taste_disturbances")

## salt
data_list <- list(salt = df1$salt, N = length(df1$salt))

# Create JAGS model
model <- jags.model(textConnection(model_string), data = data_list, n.chains = 2)

# Burn-in
update(model, 1000)

# Sampling
Salt_samples <- coda.samples(model, variable.names = c("mu", "tau"), n.iter = 5000)

## Sugar_sweeteners

data_list <- list(Sugar_samples = df1$sugar_sweeteners, N = length(df1$sugar_sweeteners))

# Create JAGS model
model <- jags.model(textConnection(model_string), data = data_list, n.chains = 2)

# Burn-in
update(model, 1000)

# Sampling
Sugar_samples <- coda.samples(model, variable.names = c("mu", "tau"), n.iter = 5000)

# View results
print(summary(Sugar_samples))
```

# Bayesian model
```{r Bayesian model, echo=TRUE, message=FALSE, warning=FALSE }
library(rstan)

model_code <- "
data {
  int<lower=0> N; // number of observations
  vector[N] y; // observations
}
parameters {
  real mu; // mean
  real<lower=0> sigma; // standard deviation
}
model {
  y ~ normal(mu, sigma); // likelihood
  mu ~ normal(0, 10); // prior for mu
  sigma ~ cauchy(0, 5); // prior for sigma
}
"

##salt
data_list <- list(y = df1$salt, N = length(df1$salt))
```


```{r Stan model , echo=TRUE, message=FALSE, warning=FALSE , results='hide'}
# Compile the Stan model
stan_model <- stan_model(model_code = model_code)
```


```{r Baysian }
# Fit the model
fit <- sampling(stan_model, data = data_list, iter = 2000, chains = 4)

traceplot(fit)


# Example summary based on hypothetical output
summary <- summary(fit)
posterior_mu <- summary$summary["mu", ]
posterior_sigma <- summary$summary["sigma", ]

cat("Posterior Mean (mu):", posterior_mu["mean"], "\n")
cat("95% Credible Interval for mu:", posterior_mu["2.5%"], "-", posterior_mu["97.5%"], "\n")

cat("Posterior Standard Deviation (sigma):", posterior_sigma["mean"], "\n")
cat("95% Credible Interval for sigma:", posterior_sigma["2.5%"], "-", posterior_sigma["97.5%"], "\n")

##f the credible interval for mu is narrow and doesn’t include 0, we might conclude that the mean salt concentration differs significantly from 0.

#Extract Posterior Samples for Further Analysis
posterior_samples <- extract(fit)
posterior_mu <- posterior_samples$mu
posterior_sigma <- posterior_samples$sigma

# Summarize the posterior samples
summary(posterior_mu)
summary(posterior_sigma)

#Posterior Predictive Checks

posterior_samples <- extract(fit)
posterior_mu <- posterior_samples$mu
posterior_sigma <- posterior_samples$sigma

# Number of posterior draws
n_draws <- 100  # Number of draws to generate

# Generate posterior predictive samples
y_rep <- matrix(NA, nrow = n_draws, ncol = length(df1$salt))

for (i in 1:n_draws) {
  # Draw new data from the normal distribution using mu and sigma
  y_rep[i, ] <- rnorm(length(df1$salt), mean = posterior_mu[i], sd = posterior_sigma[i])
}

# View the first few rows of the simulated data
head(y_rep)

# Plot observed data
hist(df1$salt, main = "Observed Salt", xlab = "Salt", col = "gray")

# Plot the first set of posterior predictive samples
hist(y_rep[1, ], main = "Simulated Salt (Posterior Predictive)", xlab = "Salt", col = "blue")

```

# permutation tests
```{r permutation tests}
library(coin)

df1$group<-factor(df1$group)
permutation_test_salt <- oneway_test(salt ~ group, data = df1, distribution = approximate(nresample = 9999))

# View results
print(permutation_test_salt)

permutation_test_sugar <- oneway_test(sugar_sweeteners~ group, data = df1, distribution = approximate(nresample = 9999))
# View results
print(permutation_test_sugar)
```

